---
title: "Zoning Cleaning"
author: "Margo La Clair"
output: html_document
date: "2023-10-28"
---

Libraries Needed:
```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(readxl)
library(sf)
library(sp)
library(ggplot2)
```

Bring in the tracked zoning spreadsheet:
```{r}
zoning_changes = read_csv("Zoning Tracker.csv", show_col_types = FALSE)
```

Find a good key (for later, when going from long to wide)!

```{r}
dups_zoning <- zoning_changes |>
  count(time_status, legislative_number_policy_name) |> #Some cities had multiple policies that went into effect on the same day, so this is a good key for later. For merging datasets later on, we will do so by municipality_name
  filter(n>1) 
stopifnot(nrow(dups_zoning) == 0)
```

The data looks pretty darn clean, except for some issues that might cause future merging issues (accents) and splitting up the reform type into multiple columns/or making it tidy if needed.  

```{r}
str(zoning_changes)
```
```{r}
#Getting rid of the accent
zoning_changes_clean <- zoning_changes |>
  mutate(municipality_name = str_replace_all(municipality_name,"é","e"))
```

```{r}
#Turning into tidy format, if needed
zoning_changes_tidy <- zoning_changes_clean |>
  separate_rows(reform_type, sep = ",") |>
  mutate(reform_type = str_squish(reform_type))
```

If I need to pivot wider instead of longer for merging purposes later on, the code for that is the following:
```{r}
zoning_changes_wide <- zoning_changes_clean |>
  separate_longer_delim(reform_type, delim = ",") |>
  group_by(time_status, legislative_number_policy_name) |>
  pivot_wider(names_from = reform_type, values_from = reform_type) |> #That did something funky with multiple columns, so fixing that
  unite(col = "TOD Reform", c("TOD Reform", "TOD Reform")) |>
  unite(col = "Plex Reform", c("Plex Reform", "Plex Reform")) |>
  unite(col = "Other Reform", c("Other Reform", "Other Reform"))
  
```

And....there are dummy variables for these already! :/ Keeping above code for posterity. 

There is quite a bit of summary information about each proposed/enacted legislation, and I want to do just a quick read-through to make sure that these policies are not introducing/expanding SFR zoning.  

```{r}
summary <- zoning_changes |>
  select(description)

write_excel_csv2(summary, file = "Policy Descriptions.csv") #Easiest way to do a visual scan. 
```

Making a quick visual to make sure the lat/long information is sound.

```{r}
sf_test <- st_as_sf(zoning_changes_clean, coords = c("longitude", "latitude"))
sp::CRS(SRS_string='EPSG:4326')
ggplot(sf_test) +
  geom_sf(aes(color=reform_phase))
```
Note that we only have data on crime statistics for 2020-2022, so I want to subset this data to only include zoning changes for years 2020 and 2021. 

```{r}
zoning_cleaned <- zoning_changes_clean |>
  filter(year == "2020" | year == "2021")
```

Saving the data file

```{r}
saveRDS(zoning_cleaned, "Zoning_Cleaned.RDA")
```



